vegan_analysis

### 2023-11-08 scripts/0_whole_game.qmd
final comments

if (!file.exists(vlce_loc)) {
  vlce <- rast("F:/mosaiced/VLCE2.0/LC_Class_HMM_2015_v20_v20.dat")
  bcb <- bcmaps::bc_bound() %>%
    vect()

  bcb_rast <- bcb %>%
    rasterize(y = vlce) %>%
    crop(bcb)

@@

  writeRaster(vlce, vlce_loc, filetype = "envi")
}
vlce <- rast(vlce_loc)
bcb <- bcmaps::bc_bound() %>%
  vect()

>>>>

bcb <- bcmaps::bc_bound_hres() %>%
    vect()

if (!file.exists(vlce_loc)) {
  vlce <- rast("F:/mosaiced/VLCE2.0/LC_Class_HMM_2015_v20_v20.dat")

  bcb_rast <- bcb %>%
    rasterize(y = vlce)

@@

  writeRaster(vlce, vlce_loc, filetype = "envi")
}
vlce <- rast(vlce_loc)

### 2023-11-08 scripts/4_vegan.qmd
final comments

```{r packages}
library(vegan)
library(tidyverse)
library(cols4all)
library(terra)
library(sf)
library(patchwork)
library(ggradar)
source(here::here("scripts", "get_keys.R"))
```

@@

@@

fracts_bar_plot <- fracts_bar_tab %>%
  ggplot(aes(x = value, y = fct_reorder(zone_nm, total), fill = str_to_title(name))) +
  geom_col(position = "stack") +
  facet_grid(rows = vars(fct_reorder(group, total, .desc = T)),
             scales = "free_y",
             space = "free") +
  labs(x = "Variance Explained",
       y = NULL,
       fill = NULL) +
  theme(legend.position = "bottom",
        panel.spacing.y = unit(0, "lines")) +
  guides(fill = guide_legend(reverse = T)) +
  scale_x_continuous(labels = scales::label_percent())

@@

##### run to here for analysis to be done

@@

>>>>

```{r packages}
library(vegan)
library(tidyverse)
library(cols4all)
library(terra)
library(sf)
library(patchwork)
# library(ggradar)
source(here::here("scripts", "get_keys.R"))
```

@@

```{r}
# linear relationships * homoschedasticity check

bec_normal %>%
  pivot_longer(cols = ends_with("DHI"),
               names_to = "dhi",
               values_to = "dhi_val") %>%
  pivot_longer(
    cols = basal_area:total_biomass,
    names_to = "struct",
    values_to = "struct_val"
  ) %>%
  ggplot(aes(x = struct_val, y = dhi_val)) +
  geom_hex() +
  facet_grid(rows = vars(dhi),
             cols = vars(struct),
             scales = "free") +
  labs(x = "Structure Value (Normalized)",
       y = "DHI Value (Normalized)")

strata_nested_long <- strata_nested %>%
  unnest(cols = data) %>%
  pivot_longer(cols = ends_with("DHI"),
               names_to = "dhi",
               values_to = "dhi_val") %>%
  pivot_longer(
    cols = basal_area:total_biomass,
    names_to = "struct",
    values_to = "struct_val"
  ) %>%
  group_by(strata) %>%
  group_split()

test <- function(df) {
  strata <- df %>%
    pull(strata) %>%
    unique()

  plot <- df %>% ggplot(aes(x = struct_val, y = dhi_val)) +
    geom_hex() +
    facet_grid(rows = vars(dhi),
               cols = vars(struct),
               scales = "free") +
    labs(x = "Structure Value (Normalized)",
         y = "DHI Value (Normalized)")

  ggsave(here::here("outputs", "linear", glue::glue("{strata}.png")))
}

map(strata_nested_long, test)
```

@@

fracts_bar_plot <- fracts_bar_tab %>%
  ggplot(aes(x = value, y = fct_reorder(zone_nm, total), fill = str_to_title(name))) +
  geom_col(position = "stack", col = "white") +
  facet_grid(rows = vars(fct_reorder(group, total, .desc = T)),
             scales = "free_y",
             space = "free") +
  labs(x = "Variance Explained",
       y = NULL,
       fill = NULL) +
  theme(legend.position = "bottom",
        panel.spacing.y = unit(0, "lines")) +
  guides(fill = guide_legend(reverse = T)) +
  scale_x_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = c("#6666ff", "#c22966", "#ff6666"))

@@

# run to here for analysis to be done

@@

# parallel coordinate pos neg

```{r}
neg_pos <- clean_loads_raw %>%
  left_join(keys$bec %>%
              select(strata = zone, group)) %>%
  left_join(keys$continuous %>%
              select(var = variable, var_long)) %>%
  mutate(group = ifelse(is.na(group), strata, group),
         var_long = ifelse(is.na(var_long), var, var_long)) %>%
  group_by(group, var_long, axis) %>%
  summarize(n_pos = sum(ifelse(axis_loading > 0, 1, 0)),
            n_neg = sum(ifelse(axis_loading < 0, 1, 0)),
            axis_loading = mean(axis_loading)
            ) %>%
  ungroup() %>%
  complete(group, var_long, axis, fill = list(axis_loading = NA,
                                              n_pos = NA,
                                              n_neg = NA)) %>%
  mutate(pos_neg = case_when(n_pos > 0 & n_neg > 0 ~ "Mixed",
                             n_pos > 0 ~ "Positive",
                             n_neg > 0 ~ "Negative")) %>%
  select(-n_pos, -n_neg, -axis_loading)

data <- clean_loads_abs %>%
  left_join(keys$bec %>%
              select(strata = zone, group)) %>%
  left_join(keys$continuous %>%
              select(var = variable, var_long)) %>%
  mutate(group = ifelse(is.na(group), strata, group),
         var_long = ifelse(is.na(var_long), var, var_long)) %>%
  group_by(group, var_long, axis) %>%
  summarize(axis_loading = mean(axis_loading)) %>%
  ungroup() %>%
  complete(group, var_long, axis, fill = list(axis_loading = NA)) %>%
  left_join(neg_pos) %>%
  mutate(inout = ifelse(endsWith(var_long, "DHI"), "Response", "Predictor"), 
         group = fct_relevel(group, "British Columbia", 
                             "Broadleaf", 
                             "Coniferous", 
                             "Mixed Wood", 
                             "Wetland-Treed", 
                             after = 0),
         var_long = fct_relevel(var_long, 
                                c("Canopy Height", "Canopy Cover", "Structural Complexity",
                                "Basal Area", "Aboveground Biomass", "Gross Stem Volume",
                                "Cumulative DHI", "Variation DHI", "Minimum DHI")))

  geom_line(aes(col = group, group = group), lty = "dashed") +
  geom_point(data = data %>% 
               filter(pos_neg == "Mixed"), aes(group = group, shape = group), size = 2.5, col = "black", fill = "black") +
  geom_point(data = data %>% 
               filter(pos_neg == "Positive"), aes(group = group, shape = group), size = 2.5, col = "red", fill = "red") +
  geom_point(data = data %>% 
               filter(pos_neg == "Negative"), aes(group = group, shape = group), size = 2.5, col = "grey", fill = "grey") +
  geom_point(aes(col = group, group = group, shape = group, fill = group)) +
  # geom_hline(yintercept = 0, lty = "dashed", col = "red") +
  facet_wrap(~paste(axis, inout),
             scale = "free") +
  #theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL,
       y = "Axis Loading (absolute)",
       col = NULL,
       shape = NULL,
       fill = NULL) +
  # coord_flip() +
  scale_colour_discrete_c4a_cat(palette = "bold") +
  scale_fill_discrete_c4a_cat(palette = "bold") +
  scale_shape_manual(values = c(21, 22, 22, 22, 22, 25, 25, 25, 25, 25))

ggsave(here::here("outputs", "parcoord_test.png"), plot = parcoord_noflip,
       height = 6, width = 8)

ggsave(here::here("outputs", "parcoord_noflip.svg"), plot = parcoord_noflip,
       height = 6, width = 8)

```{r}
data %>%
  ggplot(aes(x = var_long, y = axis_loading)) +
  geom_point(aes(shape = group, col = pos_neg, fill = group)) +
  geom_line(aes(col = group)) +
  facet_wrap(~paste(axis, inout), scales = "free") +
  scale_shape_manual(values = c(21, 22, 22, 22, 22, 25, 25, 25, 25, 25)) +
  scale_fill_discrete_c4a_cat(palette = "bold") +
  scale_colour_manual(values = c("black", "grey", "red"))

ggsave(here::here("outputs", "parcoord_test.png"), height = 6, width = 8)
```

```{r}
clean_loads_raw %>%
  mutate(pos_neg = ifelse(axis_loading > 0, "Positive", "Negative")) %>%
  count(var, axis, pos_neg) %>%
  ggplot(aes(y = var, x = n, fill = pos_neg)) +
  geom_col() +
  facet_wrap( ~ axis)

clean_loads_raw %>%
  left_join(keys$bec %>%
              select(strata = zone, group)) %>%
  mutate(pos_neg = ifelse(axis_loading > 0, "Positive", "Negative")) %>%
  count(var, group, axis, pos_neg) %>%
  mutate(n = ifelse(pos_neg == "Positive", n, -n)) %>%
  ggplot(aes(y = var, x = n, fill = pos_neg)) +
  geom_col() +
  facet_grid(cols = vars(axis),
             rows = vars(group))
```

```{r}
groups <- keys$bec %>%
  pull(group) %>% unique()

lollipop_data <- clean_loads_raw %>%
  left_join(keys$bec %>%
              select(strata = zone, group)) %>%
  left_join(keys$continuous %>%
              select(var = variable, var_long)) %>%
  mutate(
    group = ifelse(is.na(group), strata, group),
    var_long = ifelse(is.na(var_long), var, var_long),
    pos_neg = ifelse(axis_loading > 0, "Positive", "Negative")
  ) %>%
  group_by(var_long, axis, group, pos_neg) %>%
  summarize(n = n(),
            avg_loading = mean(axis_loading)) %>%
  ungroup() %>%
  mutate(
    inout = ifelse(endsWith(var_long, "DHI"), "Response", "Predictor"),
    group = fct_relevel(
      group,
      "British Columbia",
      "Broadleaf",
      "Coniferous",
      "Mixed Wood",
      "Wetland-Treed",
      after = 0
    ),
    var_long = fct_relevel(
      var_long,
      c(
        "Canopy Height",
        "Canopy Cover",
        "Structural Complexity",
        "Basal Area",
        "Aboveground Biomass",
        "Gross Stem Volume",
        "Cumulative DHI",
        "Variation DHI",
        "Minimum DHI"
      )
    )
  )

lollipops <- lollipop_data %>%
  group_by(group %in% groups) %>%
  group_split()

left <- lollipops[[1]] %>%
  ggplot(aes(x = avg_loading,
             y = var_long,
             col = pos_neg)) +
  geom_point() +
  geom_segment(aes(
    x = avg_loading,
    xend = 0,
    y = var_long,
    yend = var_long
  )) +
  geom_vline(xintercept = 0,
             alpha = 0.5,
             lty = "dotted") +
  facet_grid(cols = vars(axis),
             rows = vars(group)) +
  theme(legend.position = "bottom") +
  labs(x = "Average Loading",
       y = NULL)

right = lollipops[[2]] %>%
  ggplot(aes(x = avg_loading,
             y = var_long,
             col = pos_neg)) +
  geom_point(aes(size = n)) +
  geom_segment(aes(
    x = avg_loading,
    xend = 0,
    y = var_long,
    yend = var_long
  )) +
  geom_vline(xintercept = 0,
             alpha = 0.5,
             lty = "dotted") +
  facet_grid(cols = vars(axis),
             rows = vars(group)) +
  theme(legend.position = "bottom",
        axis.text.y = element_blank()) +
  labs(x = "Average Loading",
       y = NULL)

left + right

ggsave(here::here("outputs", "chris.png"))
```

```{r}
lollipop_data %>%
  ggplot(aes(x = var_long, y = avg_loading)) +
  #geom_boxplot(col = "#00000025", outlier.alpha = 0) +
  geom_line(data = lollipop_data %>%
              filter(pos_neg == "Positive"), aes(col = group, group = group), lty = "dashed") +
  geom_line(data = lollipop_data %>%
              filter(pos_neg == "Negative"), aes(col = group, group = group), lty = "dashed") +
  geom_point(aes(col = group, group = group, shape = group, fill = group)) +
  # geom_hline(yintercept = 0, lty = "dashed", col = "red") +
  facet_wrap(~paste(axis, inout),
             scale = "free") +
  #theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL,
       y = "Axis Loading (absolute)",
       col = NULL,
       shape = NULL,
       fill = NULL) +
  # coord_flip() +
  scale_colour_discrete_c4a_cat(palette = "bold") +
  scale_fill_discrete_c4a_cat(palette = "bold") +
  scale_shape_manual(values = c(21, 22, 22, 22, 22, 25, 25, 25, 25, 25))
```

# flip cum to positive

```{r}
zones <- keys$bec %>% pull(zone)

flip_cum <- function(data) {
  if(data %>% filter(var == "CumDHI") %>% nrow() == 0) {
    return(data)
  }
  flip <- data %>%
    filter(var == "CumDHI") %>%
    pull(axis_loading) %>%
    sign()

  data %>% mutate(axis_loading = axis_loading * flip)
}

clean_loads_cum_pos <- clean_loads_raw %>%
  group_by(strata, axis) %>%
  group_split() %>%
  map_dfr(., .f = flip_cum)

clean_loads_cum_pos %>%
  filter(strata %in% zones) %>%
  count(var, axis, sign = sign(axis_loading)) %>%
  pivot_wider(names_from = sign, values_from = n) %>%
  arrange(axis)

```{r}
groups <- keys$bec %>% pull(group) %>% unique()

clean_loads_joined <- clean_loads_cum_pos %>%
  left_join(keys$bec %>%
              select(strata = zone, group)) %>%
  left_join(keys$continuous %>%
              select(var = variable, var_long)) %>%
  mutate(
    group = ifelse(is.na(group), strata, group),
    var_long = ifelse(is.na(var_long), var, var_long),
    pos_neg = ifelse(axis_loading > 0, "Positive", "Negative")
  )

lollipop_data <- clean_loads_joined %>%
  group_by(var_long, axis, group, pos_neg) %>%
  summarize(n = n(),
            avg_loading = mean(axis_loading)) %>%
  ungroup() %>%
  mutate(
    inout = ifelse(endsWith(var_long, "DHI"), "Response", "Predictor"),
    group = fct_relevel(
      group,
      "British Columbia",
      "Broadleaf",
      "Coniferous",
      "Mixed Wood",
      "Wetland-Treed",
      after = 0
    ),
    var_long = fct_relevel(
      var_long,
      c(
        "Canopy Height",
        "Canopy Cover",
        "Structural Complexity",
        "Basal Area",
        "Aboveground Biomass",
        "Gross Stem Volume",
        "Minimum DHI",
        "Variation DHI",
        "Cumulative DHI"
      )
    )
  )

lollipops <- lollipop_data %>%
  group_by(group %in% groups) %>%
  group_split()

left <- lollipops[[1]] %>%
  ggplot(aes(x = avg_loading,
             y = var_long,
             col = pos_neg)) +
  geom_point() +
  geom_segment(aes(
    x = avg_loading,
    xend = 0,
    y = var_long,
    yend = var_long
  )) +
  geom_vline(xintercept = 0,
             alpha = 0.5,
             lty = "dotted") +
  facet_grid(cols = vars(axis),
             rows = vars(group)) +
  theme(legend.position = "bottom") +
  labs(x = "Average Loading",
       y = NULL)

right = lollipops[[2]] %>%
  ggplot(aes(x = avg_loading,
             y = var_long,
             col = pos_neg)) +
  geom_point(aes(size = n)) +
  geom_segment(aes(
    x = avg_loading,
    xend = 0,
    y = var_long,
    yend = var_long
  )) +
  geom_vline(xintercept = 0,
             alpha = 0.5,
             lty = "dotted") +
  facet_grid(cols = vars(axis),
             rows = vars(group)) +
  theme(legend.position = "bottom",
        axis.text.y = element_blank()) +
  labs(x = "Average Loading",
       y = NULL)

left + right

ggsave(here::here("outputs", "chris.png"))
```

```{r}
data <- clean_loads_joined %>%
  # mutate(axis_loading = abs(axis_loading)) %>%
  group_by(group, var_long, axis) %>%
  summarize(axis_loading = mean(axis_loading)) %>%
  ungroup() %>%
  complete(group, var_long, axis, fill = list(axis_loading = NA)) %>%
  mutate(inout = ifelse(endsWith(var_long, "DHI"), "Response", "Predictor"), 
         group = fct_relevel(group, "British Columbia", 
                             "Broadleaf", 
                             "Coniferous", 
                             "Mixed Wood", 
                             "Wetland-Treed", 
                             after = 0),
         var_long = ifelse(endsWith(var_long, "DHI"), var_long, str_replace_all(var_long, " ", "\n")),
         var_long = fct_relevel(var_long, 
                                c("Structural\nComplexity", "Canopy\nHeight", "Canopy\nCover",
                                "Basal\nArea", "Aboveground\nBiomass", "Gross\nStem\nVolume",
                                "Cumulative DHI", "Variation DHI", "Minimum DHI")))

  geom_line(aes(col = group, group = group), lty = "dashed") +
  geom_point(aes(col = group, group = group, shape = group, fill = group)) +
  geom_hline(yintercept = 0, lty = "dotted", col = "grey", linewidth = 0.5) +
  facet_wrap(~paste(axis, inout),
             scale = "free") +
  #theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL,
       y = "Axis Loading",
       col = NULL,
       shape = NULL,
       fill = NULL) +
  # coord_flip() +
  scale_colour_discrete_c4a_cat(palette = "bold") +
  scale_fill_discrete_c4a_cat(palette = "bold") +
  scale_shape_manual(values = c(21, 22, 22, 22, 22, 25, 25, 25, 25, 25))

# ggsave(here::here("outputs", "parcoord_noflip.svg"), plot = parcoord_noflip,
#        height = 6, width = 8)
```

# meg scatterplots
```{r}
scatter <- bec_normal %>%
  mutate(label = case_when(bec %in% c("CWH", "MS") ~ bec,
                           T ~ "Other"))

ggplot(data = scatter, aes(col = label), alpha = 0.25) +
  geom_point(aes(x = percentage_first_returns_above_2m, y = CumDHI)) +
  geom_density(aes(x = percentage_first_returns_above_2m)) +
  geom_density(aes(y = CumDHI)) +
  scale_fill_manual(values = c("grey", "red", "blue")) +
  scale_colour_manual(values = c("grey", "red", "blue"))

ggplot() +
  geom_point(data = bec_df, aes(x = percentage_first_returns_above_2m, y = CumDHI), alpha = 0.5) +
  geom_point(data = bec_df %>%
               filter(bec == "CWH"), aes(x = percentage_first_returns_above_2m, y = CumDHI), col = "orange") +
  geom_point(data = bec_df %>%
               filter(bec == "MS"), aes(x = percentage_first_returns_above_2m, y = CumDHI), col = "red") +
  geom_histogram(data = bec_df, aes(x = percentage_first_returns_above_2m)) +
  geom_histogram(data = bec_df, aes(y = CumDHI))
```

```{r}
fracts_out

strata_nested %>% 
  unnest(cols = data) %>%
  select(-bec, -category) %>%
  left_join(keys$bec, by = c("strata" = "zone")) %>%
  mutate(group = ifelse(is.na(group), "Forest Type", group)) %>%
  ggplot() +
  geom_boxplot(aes(x = fct_reorder(.f = strata, .x = MinDHI, .fun = mean), y = MinDHI, fill = group))
```

```{r}
strata_nested %>% 
  unnest(cols = data) %>%
  select(-bec, -category) %>%
  group_by(strata) %>%
  summarize(across(basal_area:VarDHI, .fns = sd)) %>%
  left_join(keys$bec, by = c("strata" = "zone")) %>%
  mutate(group = ifelse(is.na(group), "Forest Type", group)) %>%
  select(strata, group, basal_area:VarDHI) %>%
  pivot_longer(-c(strata, group)) %>%
  left_join(fracts_out) %>%
  mutate(varexp = primary + modelled + overlap) %>%
  ggplot(aes(x = value, y = varexp, col = group)) +
  geom_point() +
  facet_wrap(~name, strip.position = "bottom") +
  labs(x = NULL,
       y = "Variation Explained",
       title = "x axis is sd of variable") +
  theme(strip.placement = "outside",
        strip.background = element_blank())
```

```{r}
bec_df %>%
  left_join(fracts_out %>%
              mutate(all = (primary + modelled + overlap) * 100) %>%
              select(strata, all), by = c("bec" = "strata")) %>%
  mutate(facet = paste0(bec, " (", round(all, 2), "%)")) %>%
  ggplot(aes(x = percentage_first_returns_above_2m / 100, y = CumDHI)) +
  geom_hex(aes(fill = after_stat(log(count)))) +
  labs(x = "Canopy Cover (%)") +
  facet_wrap(~fct_reorder(facet, all)) +
  theme(legend.position = "bottom")

ggsave(here::here("outputs", "supplement_hex.png"),
       height = 6, width = 8)
```
